---
title: 'Assignment 3 Task 4: Naples Reef (2010) purple urchins'
author: "Karina Johnston"
date: "10/27/2021"
output: 
  html_document: 
    toc: yes
    theme: cerulean
---

This assignment will analyze sea urchin data from the Santa Barbara Coastal Long Term Ecological Research program. Data for purple (*Strongylocentrotus purpuratus*) urchins observed at Naples Reef along the Santa Barbara coast are analyzed. Exploratory figures and a finalized summary table are produced, exploring data visualization techniques and summary statistics. 

Citation: Reed, D. 2020. SBC LTER: Reef: Long-term experiment: Kelp removal: Urchin size frequency distribution ver 20. Environmental Data Initiative. Accessed 2010-10-27. (https://doi.org/10.6073/pasta/5a1e9ef03aa47bd2225c0bb98a02a63b)

```{r setup, include = TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Attach the following packages into the setup code chunk and change the settings for reporting warnings and messages (above)
library(tidyverse)
library(kableExtra)
library(here)
library(janitor)
library(lubridate)
```

### Read in the data

This step reads in the urchin data .csv file from within the project folder and creates a new dataframe `urchin_data`, followed by a step to change the column headers to lower snake case using the function `clean_names()` found in the `janitor` package.

```{r}
urchin_data <- read_csv(here("data", "urchins.csv")) %>% 
  clean_names()
```

### Section 1: Data wrangling

First, the urchins data are filtered to only include Naples Reef data for purple urchins in 2010 and saved to a new dataframe `purple_naples_2010`.

```{r}
# creates the new dataframe beginning with urchin_data
purple_naples_2010 <- urchin_data %>% 

# updates the date column to be Date format and adds a year column 
  mutate(date = mdy(date)) %>% 
  mutate(year = year(date)) %>% 

# filters only purple urchins from 2010 and Naples Reef
  filter(common_name == "Purple Urchin", 
         site == "NAPL",
         year == "2010") %>% 
  
```



