---
title: "Assignment 3 Task 3 - Santa Barbara LTER Urchins"
author: "Karina Johnston"
date: "10/27/2021"
output: 
  html_document: 
    theme: cerulean
---

This assignment will analyze sea urchin data from the Santa Barbara Coastal Long Term Ecological Research program. Data for both red (*Mesocentrotus franciscanus*) and purple (*Strongylocentrotus purpuratus*) urchins observed at five sites along the Santa Barbara coast are analyzed. 

Citation: Reed, D. 2020. SBC LTER: Reef: Long-term experiment: Kelp removal: Urchin size frequency distribution ver 20. Environmental Data Initiative. Accessed 2010-10-27. (https://doi.org/10.6073/pasta/5a1e9ef03aa47bd2225c0bb98a02a63b)

```{r setup, include = TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Attach the packages into the setup code chunk
library(tidyverse)
library(kableExtra)
library(here)
library(janitor)
```

### Read in the data

This step reads in the urchin data .csv file from within the project folder and creates a new dataframe `urchin_data`, followed by a step to change the column headers to lower snake case using the function `clean_names()` found in the `janitor` package.

```{r}
urchin_data <- read_csv(here("data", "urchins.csv")) %>% 
  clean_names()
```

### Section 1: Size Comparisons

In this section, the data will be visualized first by comparing the distributions of sizes between purple and red urchins with a finalized, faceted histogram using the `ggplot()` function with cleaned up aesthetics. 

```{r, fig.cap = "Figure 1. Size frequency data for Santa Barbara coastal red and purple urchins (data: SBC LTER)."}
ggplot(data = urchin_data, aes(x = size)) +
  geom_histogram(bins = 15, color = 4, fill = 4) +
  facet_wrap(~ common_name) +
  labs(x = "Size (cm)", y = "Number of urchins", title = "Size Frequency Histogram for Santa Barbara Urchins (Johnston)") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        panel.grid = element_blank())
```

Next, a finalized summary table is created that contains the following statistics for purple and red urchins (mean, standard deviation or SD, and sample size).

```{r}

```

